### 2019-08-12 manuscript.Rmd
manuscript

To start, the users need to install \pkg{GNU MCSim} [@JSSv002i09] and related standard \proglang{C} compiler to solve the ODE function. In addition, \pkg{pksensi} can link with \proglang{R} \pkg{deSolve} package, which includes a comprehensive integrator to solve ODE. The \pkg{GNU MCSim} can be installed by following the instruction in \pkg{GNU MCSim}’s manual on https://www.gnu.org/software/mcsim/mcsim.html or using the built-in function \code{mcsim_install} in \pkg{pksensi}. The \proglang{C} compiler is already available in Linux or MacOS. The Windows user can install Rtools on https://cran.r-project.org/bin/windows/Rtools/ and set the working path of compiler (\code{/Rtools/mingw_64/bin} and \code{/Rtools/bin}) before using this function as:

@@

To start, we implemented the one-compartment PK model in \proglang{R} by compiling the file written under the \proglang{C} or \pkg{GNU MCSim} format. \pkg{pksensi} allows users select the preferred method to solve the PK model, either with the \pkg{deSolve} package or with \pkg{GNU MCSim} through the compile function. Running model under \pkg{GNU MCSim} native code can have a faster speed to obtain the model outputs. This section mainly focus on how to conduct global SA with \pkg{deSolve} package. Then compare the computing time with \pkg{GNU MCSim}.

The following \proglang{R} script will download the one-compartment PK model from https://github.com/nanhung/pksensi/blob/master/tests/pbtk1cpt.model. The model was developed on \pkg{GNU MCSim}'s format that mainly includes two state variables, which are the quantity of compound in the gut lumen (\code{Agutlument}) and the rest of body (\code{Acompartmant}). The \code{Ametabolized} is the quantity of compound transform and metabolize through hepatic clearance. The \code{AUC} is the calculated area under the curve of the rest body ($mg \cdot h/L$).

The code must first be compiled to run the model. After create the model file, we can use \code{compile_model} to generate the file that has dynamic-link library (.dll) or share object (.so) extension and can be linked dynamically into an \proglang{R} session (\code{pbtk1cpt.dll} on Windows or \code{pbtk1cpt.so} on other systems) and \proglang{R} file (\code{pbtk1cpt_inits.R}) with default input parameters and initial state settings with the definition of \code{application = "R"}.

```{r}
mName <- "pbtk1cpt"
```

```{r eval=T, message=FALSE, warning=FALSE}
compile_model(mName, application = "R")
```

The \code{pbtk1cpt_inits.R} file includes \code{initParms} and \code{initStates} functions and \code{Outputs} variable. The created function have default value of model parameters and initial conditions that can further use to customize in the simulation.

```{r, eval = T}
source(paste0(mName, "_inits.R"))
```

On the other side, if the user only has the \proglang{C} version of the model code (\code{pbtk1cpt.c}), it can be compiled and loaded as:

```{r, eval = F}
system("R CMD SHLIB pbtk1cpt.c")
dyn.load(paste0("pbtk1cpt", .Platform$dynlib.ext))
```

The parameter values and initial states can be customized to the specific condition. It can also schedule for the given dosing scenario. In the current setting, we assumed the initial condition of the intake dose to be 10 mg. We can use \code{initParms} and \code{initStates} functions to customize the parameter values and the initial state that will be used in the following modeling and SA. These additional functions are generated when compiling the model file through \pkg{GNU MCSim}. Therefore, if the model was compiled through \code{R CMD SHLIB}, the user need to assign the baseline value of model parameters and state variable on their own.

We used the corresponding parameter value of acetaminophen (APAP) in this example. These model parameters are derived from the *in-vivo* or *in-vitro* experiment results. The parameter value can be generated from \code{parameterize_1comp} function in \pkg{httk} package as:

@@

```{r, eval=T}
parms <- initParms()
parms["vdist"] <- 1.1
parms["ke"] <- 0.23
parms["kgutabs"] <- 2.18
parms["BW"] <- 70
initState <- initStates(parms=parms)
initState["Agutlument"] <- 10
```

Here shows the given parameter value (\code{parms}), initial state condition (\code{initState}), and the output variable (\code{Outputs}) that need to specify in model solving.

```{r, eval=T}
parms
```

```{r, eval=T}
initState
```

Both \code{parms} and \code{initState} are "numeric variables" that contain the value of parameter and initial state condition.

```{r, eval=T}
Outputs
```

The \code{Outputs} is a "character" that specifies the estimated model output of concentration in rest body.

@@

```{r, eval=T}
t <- seq(from = 0.01, to = 24.01, by = 1)
y <- deSolve::ode(initState, t, func = "derivs", parms = parms, 
         dllname = mName, initfunc = "initmod", 
         nout = 1, outnames = Outputs)
```

@@

```{r, eval=T}
q <- c("qunif", "qunif", "qunif", "qnorm")
q.arg <- list(list(min = parms["vdist"] / 2, max = parms["vdist"] * 2),
             list(min = parms["ke"] / 2, max = parms["ke"] * 2), 
             list(min = parms["kgutabs"] / 2, max = parms["kgutabs"] * 2),
             list(mean = parms["BW"], sd = 5)) 
params <- c("vdist", "ke", "kgutabs", "BW")
```

@@

```{r, eval=T}
set.seed(1234)
x <- rfast99(params, n = 200, q = q, q.arg = q.arg, replicate = 20)
```

@@

Because the PK model is being used to describe a continuous process for the chemical concentration over time, the sensitivity measurements, therefore, have the time-dependent relationships for each model parameter. Here we define the output time points (\code{t}) to examine the change of the parameter sensitivity over time. To solve the model through \pkg{deSolve}, we need to provide the details of the argument, which include time (\code{t}), initial  conditions of state variable (\code{initState}), output variables (\code{outnames}), and name of the shared library (\code{mName}, without extension). To create the time-dependent sensitivity measurement, we set the time duration from 0.01 to 24.01 hours with the time segment of 1 hour as the above definition in \code{ode} function in this example. The initial time point should avoid 0 to prevent computational error in misconduct. The \code{outnames}, \code{dllname}, are based on the arguments from the \code{ode} function in \pkg{deSolve} package. The details of the model structure and these arguments are defined in \code{pbtk1comp.c}. and \code{pbtk1comp_inits.R} files.

```{r, eval=T}
outputs <- c("Ccompartment", "Ametabolized")
out <- solve_fun(x, time = t, initState = initState, 
                 outnames = outputs, dllname = mName)
```

@@

In addition to using \pkg{deSolve} to solve differential equations in PK model, the \pkg{GNU MCSim} can also be used, and provides better computational efficiency. To solve ODEs through \pkg{GNU MCSim}, we need to change the argument to \code{application = mcsim} in \code{compile_model}. The computing time of using \code{solve_fun} in SA is estimated as,

```{r, eval=T}
system.time(out<-solve_fun(x, t, initState = initState, 
                           outnames = Outputs, dllname = mName))
```

@@

```{r, eval=T, message=F, warning=F}
compile_model(mName, application = "mcsim")
```

@@

```{r, eval=T}
conditions <- c("Agutlument = 10") 
system.time(out <- solve_mcsim(x, mName = mName, params = params, 
                               vars = Outputs, time = t, 
                               condition = conditions))
```

After solving the equations under the same given condition, we can find that \pkg{GNU MCSim} has about 4 - 5 times faster in computing performance than using \pkg{deSolve}. In this case, we only focus on performing the global SA alone for generic PK model without additional comparison with experimental data. The next example will display and reproduce our previous published result [@fphar201800588] with full global SA workflow.

>>>>

In addition, \pkg{pksensi} can link with \proglang{R} \pkg{deSolve} package, which includes a comprehensive integrator to solve ODE. The \pkg{GNU MCSim} can be installed by following the instruction in \pkg{GNU MCSim}’s manual on https://www.gnu.org/software/mcsim/mcsim.html or using the built-in function \code{mcsim_install} in \pkg{pksensi}. The \proglang{C} compiler is already available in Linux or MacOS. The Windows user can install Rtools on https://cran.r-project.org/bin/windows/Rtools/ and set the working path of compiler (\code{/Rtools/mingw_64/bin} and \code{/Rtools/bin}) before using this function as:

@@

To start, we implemented the one-compartment PK model in \proglang{R}. \pkg{pksensi} allows users select the preferred method to solve the PK model, either with the \pkg{deSolve} package or with \pkg{GNU MCSim} through the compile function. Running model under \pkg{GNU MCSim} native code can have a faster speed to obtain the model outputs. This section mainly focus on how to conduct global SA with \pkg{deSolve} package with pure \proglang{R} environment. Then compare the computing time with \pkg{GNU MCSim}.

The one-compartment PK model was mainly includes two state variables, which are the quantity of compound in the gut lumen (\code{Agutlument}) and the rest of body (\code{Acompartmant}). The \code{Ametabolized} is the quantity of compound transform and metabolize through hepatic clearance. The \code{AUC} is the calculated area under the curve of the rest body ($mg \cdot h/L$).

```{r, eval=T}
pbtk1cpt <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dAgutlument = - kgutabs * Agutlument
    dAcompartment = kgutabs * Agutlument - ke * Acompartment
    dAmetabolized = ke * Acompartment
    Ccompartment = Acompartment / vdist * BW;
    list(c(dAgutlument, dAcompartment, dAmetabolized), 
         "Ccompartment" = Ccompartment) 
  })
}
```

The parameter values and initial states need to assign to the specific condition. We used the corresponding parameter value of acetaminophen (APAP) in this example. These model parameters are derived from the *in-vivo* or *in-vitro* experiment results. The parameter value can be generated from \code{parameterize_1comp} function in \pkg{httk} package as:

@@

```{r, eval=T}
parms <- c(vdist = pars1comp$Vdist, 
           ke = pars1comp$kelim, 
           kgutabs = pars1comp$kgutabs, 
           BW = pars1comp$BW)
initState <- c(Agutlument = 10, Acompartment = 0, Ametabolized = 0)
```

Here shows the given parameter value (\code{parms}) and initial state condition (\code{initState}) that need to specify in model solving. Both \code{parms} and \code{initState} are "numeric variables" that contain the value of parameter and initial state condition.

@@

```{r, eval=T}
library(deSolve)
t <- seq(from = 0.01, to = 24.01, by = 1)
y <- ode(y = initState, times = t, func = pbtk1cpt, parms = parms)
```

@@

```{r, eval=T}
q <- c("qunif", "qunif", "qunif", "qnorm")
q.arg <- list(list(min = pars1comp$Vdist / 2, max = pars1comp$Vdist * 2),
              list(min = pars1comp$kelim / 2, max = pars1comp$kelim * 2),
              list(min = pars1comp$kgutabs / 2, max = pars1comp$kgutabs * 2),
              list(mean = pars1comp$BW, sd = 5))
params <- c("vdist", "ke", "kgutabs", "BW")
```

@@

```{r, eval=T}
set.seed(1234)
x <- rfast99(params, n = 200, q = q, q.arg = q.arg, replicate = 10)
```

@@

Because the PK model is being used to describe a continuous process for the chemical concentration over time, the sensitivity measurements, therefore, have the time-dependent relationships for each model parameter. Here we use the defined output time points (\code{t}) to examine the change of the parameter sensitivity over time. To solve the model through \pkg{deSolve}, we need to provide the details of the argument, which include time (\code{t}), initial  conditions of state variable (\code{initState}), output variables (\code{outnames}), and name of the model function (\code{func}). To create the time-dependent sensitivity measurement, we set the time duration from 0.01 to 24.01 hours with the time segment of 1 hour as the above definition in \code{ode} function in this example. The initial time point should avoid 0 to prevent computational error in misconduct. The \code{outnames} is based on the arguments from the \code{ode} function in \pkg{deSolve} package.

```{r, eval=T}
outputs <- c("Ccompartment", "Ametabolized")
out <- solve_fun(x, time = t, func = pbtk1cpt, initState = initState, outnames = outputs)
```

@@

In addition to using \pkg{deSolve} to solve differential equations in PK model, the \pkg{GNU MCSim} can also be used, and provides better computational efficiency. To start, the users need to install \pkg{GNU MCSim} [@JSSv002i09] and related standard \proglang{C} compiler to solve the ODE function. To solve ODEs through \pkg{GNU MCSim}, we need to change the argument to \code{application = mcsim} in \code{compile_model}. The following \proglang{R} script will download the one-compartment PK model from https://github.com/nanhung/pksensi/blob/master/tests/pbtk1cpt.model.

The computing time of using \code{solve_fun} in SA is estimated as,

```{r, eval=T}
system.time(out <- solve_fun(x, time = t, 
                             func = pbtk1cpt, initState = initState, 
                             outnames = outputs))
```

@@

```{r, eval=T, message=F, warning=F}
mName <- "pbtk1cpt"
compile_model(mName, application = "mcsim")
```

@@

```{r, eval=T}
conditions <- c("Agutlument = 10") 
system.time(out <- solve_mcsim(x, mName = mName, params = params, 
                               vars = outputs, time = t, 
                               condition = conditions))
```

After solving the equations under the same given condition, we can find that \pkg{GNU MCSim} has more than 50 times faster in computing performance than using \pkg{deSolve}. In this case, we only focus on performing the global SA alone for generic PK model without additional comparison with experimental data. The next example will display and reproduce our previous published result [@fphar201800588] with full global SA workflow.
